### 2017-11-07 submission/manuscript.Rmd
add result edits

***Fecal Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
Using power transformed and Z-score normalized alpha diversity metrics both evenness and the Shannon diversity metrics in feces are lower in those with carcinoma then in controls but not for tissue samples [Figure 1]. Using linear mixed-effects to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. There was no increased relative risk for these metrics for adenoma or for tissue in general [Figure 2A & S1].

Using the Bray-Curtis distance metric, the fecal microbiota did not have a different community diversity between adenoma and control but did for carcinoma across studies [Table S1 & S2]. The majority of unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4]. All matched tissue samples accross studies had no difference between any of the compared groups [Table S3 & S4].

***Genera Previously Associated with Carcinoma Increases Relative Risk More than Alpha Diversity:***
Both fecal and tissue samples had a significantly increased RR for carcinoma but not for adenoma [Figure 3]. The relative risk for feces was greater than either evenness or Shannon diversity [Figure 2 & 3]. The relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The RR effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that tissue samples include matched samples.

@@

***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
For both fecal and tissue samples (matched and unmatched) there was a decrease in AUC when only OTUs from the CRC associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S2-S3]. Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S4-S5]. Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] then carcinoma models [Figure S4B & S5A].

***Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect a difference that was eqaul to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].

>>>>

***Diversity is Lower in Those with Carcinoma and Increases Relative Risk:***
Using power transformed and Z-score normalized alpha diversity metrics, both evenness and Shannon diversity in feces and not tissue were  lower in those with carcinoma [Figure 1]. Using linear mixed-effect models to control for study and variable region there was a significant decrease from control to adenoma to carcinoma for both evenness (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannoneven")[, "BH"], digits = 2, nsmall = 2))`) and Shannon diversity (P-value = `r paste(format(filter(stool_lme, alpha_metric == "shannon")[, "BH"], digits = 2, nsmall = 2))`). This effect was not observed in tissue when additionally controlling for whether the sample came from the same individual (P-value `r ifelse(min(match_tissue_lme$BH) < 0.05 & min(unmatch_tissue_lme$BH) < 0.05, invisible("check tables"), invisible(">"))` 0.05). For fecal samples a decrease in Shannon diversity and evenness resulted in a significantly increased relative risk for carcinoma (P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannon")[, "pvalue"], digits = 2, nsmall = 2))` and P-value = `r paste(format(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`, respectively) [Figure 2]. Although these values were significant, the effect size was relatively small for both metrics (Shannon RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannon")[, "rr"], digits = 2))` and evenness RR = `r paste(round(filter(alpha_RR_crc_stool, measure == "shannoneven")[, "rr"], digits = 2))`) [Figure 2]. Only a decrease in evenness had an increased relative risk for adenoma (P-value = `r paste(format(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "pvalue"], digits = 2, nsmall = 2))`) [Figure 2A & S1] but this effect size was even smaller than what was observed for carcinoma (RR = `r paste(round(filter(alpha_RR_adn_stool, measure == "shannoneven")[, "rr"], digits = 2))`).

Using the Bray-Curtis distance metric, there was a significant difference across studies in the bacterial community of fecal samples between carcinoma and controls but not between adenoma and controls [Table S1 & S2]. Many studies with unmatched tissue samples had a significant difference for both adenoma and carcinoma versus controls [Table S3 & S4] while studies with matched tissue samples had no differences [Table S3 & S4].

***Genera Previously Associated with Carcinoma Predominately Increases Relative Risk for Carcinoma but not Adenoma:***
The majority of CRC associated genera for both feces and tissue had a significantly increased relative risk for carcinoma but not for adenoma [Figure 3]. In fecal samples the relative risk due to CRC associated genera was greater than either evenness or Shannon diversity [Figure 2 & 3]. Additionally, the relative risk of carcinoma continuously increased as individuals tested positive for more CRC associated genera [Figure 3B & 3D]. The relative risk effect size was greater for stool (RR range = `r paste(round(min(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_stool, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`) then for tissue (RR range = `r paste(round(min(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))` - `r paste(round(max(filter(select_RR_tissue, !(measure %in% c("all_four", "total_four")))[, "rr"]), digits = 2))`). This decrease may be explained by the fact that the tissue analysis include matched samples.

@@

***Using the Whole Community Increases Model AUC over CRC Associated Genera:***
For both fecal and tissue samples (matched and unmatched) the AUC decreases when only OTUs from the CRC associated genera are used [Figure 4 & 5]. This decrease is observed in both adenoma and carcinoma groups [Figure 4 & 5]. The genus models generally had similar trends as observed for the OTU based models with the full genera models performing better then the CRC associated genera models [Figure S2-S3]. Both genus models perform similarily in their ability to be able to predict lesion (adenoma or carcinoma) with carcinoma having a higher AUC then adenoma [Figure S4-S5]. Matched tissue samples for those with carcinoma had an AUC that was more similar to the adenoma models [Figure S4A, S5B, & S6] then carcinoma models [Figure S4B & S5A].

***Majority of Studies are Underpowered for Detecting Small Effect Size Differences:***
When assessing the power of each study at different effect sizes the majority of studies for both adenoma and carcinoma have an 80% power to detect a 30% difference [Figure 6A & B]. No single study that was analyzed had the standard 80% power to detect an effect size difference that was eqaul to or below 10% [Figure 6A & B].  In order to achieve adequate power for small effect sizes it would be necessary to recruit over 1000 individuals for each arm of the study [Figure 6C].

### 2017-11-07 submission/manuscript.tex
add result edits

\textbf{\emph{Fecal Diversity is Lower in Those with Carcinoma and
Increases Relative Risk:}} Using power transformed and Z-score
normalized alpha diversity metrics both evenness and the Shannon
diversity metrics in feces are lower in those with carcinoma then in
controls but not for tissue samples {[}Figure 1{]}. Using linear
mixed-effects to control for study and variable region there was a
significant decrease from control to adenoma to carcinoma for both
evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
effect was not observed in tissue when additionally controlling for
whether the sample came from the same individual (P-value \textgreater{}
0.05). For fecal samples a decrease in Shannon diversity and evenness
resulted in a significantly increased relative risk for carcinoma
(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. There was no increased relative risk for these metrics
for adenoma or for tissue in general {[}Figure 2A \& S1{]}.

Using the Bray-Curtis distance metric, the fecal microbiota did not have
a different community diversity between adenoma and control but did for
carcinoma across studies {[}Table S1 \& S2{]}. The majority of unmatched
tissue samples had a significant difference for both adenoma and
carcinoma versus controls {[}Table S3 \& S4{]}. All matched tissue
samples accross studies had no difference between any of the compared
groups {[}Table S3 \& S4{]}.

\textbf{\emph{Genera Previously Associated with Carcinoma Increases
Relative Risk More than Alpha Diversity:}} Both fecal and tissue samples
had a significantly increased RR for carcinoma but not for adenoma
{[}Figure 3{]}. The relative risk for feces was greater than either
evenness or Shannon diversity {[}Figure 2 \& 3{]}. The relative risk of
carcinoma continuously increased as individuals tested positive for more
CRC associated genera {[}Figure 3B \& 3D{]}. The RR effect size was
greater for stool (RR range = 1.61 - 2.74) then for tissue (RR range =
1.21 - 1.81). This decrease may be explained by the fact that tissue
samples include matched samples.

@@

\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
Associated Genera:}} For both fecal and tissue samples (matched and
unmatched) there was a decrease in AUC when only OTUs from the CRC
associated genera are used {[}Figure 4 \& 5{]}. This decrease is
observed in both adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The
genus models generally had similar trends as observed for the OTU based
models with the full genera models performing better then the CRC
associated genera models {[}Figure S2-S3{]}. Both genus models perform
similarily in their ability to be able to predict lesion (adenoma or
carcinoma) with carcinoma having a higher AUC then adenoma {[}Figure
S4-S5{]}. Matched tissue samples for those with carcinoma had an AUC
that was more similar to the adenoma models {[}Figure S4A, S5B, \& S6{]}
then carcinoma models {[}Figure S4B \& S5A{]}.

\textbf{\emph{Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} When assessing the power of each study at
different effect sizes the majority of studies for both adenoma and
carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
B{]}. No single study that was analyzed had the standard 80\% power to
detect a difference that was eqaul to or below 10\% {[}Figure 6A \&
B{]}. In order to achieve adequate power for small effect sizes it would
be necessary to recruit over 1000 individuals for each arm of the study
{[}Figure 6C{]}.

>>>>

\textbf{\emph{Diversity is Lower in Those with Carcinoma and Increases
Relative Risk:}} Using power transformed and Z-score normalized alpha
diversity metrics, both evenness and Shannon diversity in feces and not
tissue were lower in those with carcinoma {[}Figure 1{]}. Using linear
mixed-effect models to control for study and variable region there was a
significant decrease from control to adenoma to carcinoma for both
evenness (P-value = 0.025) and Shannon diversity (P-value = 0.043). This
effect was not observed in tissue when additionally controlling for
whether the sample came from the same individual (P-value \textgreater{}
0.05). For fecal samples a decrease in Shannon diversity and evenness
resulted in a significantly increased relative risk for carcinoma
(P-value = 0.01 and P-value = 0.0011, respectively) {[}Figure 2{]}.
Although these values were significant, the effect size was relatively
small for both metrics (Shannon RR = 1.31 and evenness RR = 1.34)
{[}Figure 2{]}. Only a decrease in evenness had an increased relative
risk for adenoma (P-value = 0.032) {[}Figure 2A \& S1{]} but this effect
size was even smaller than what was observed for carcinoma (RR = 1.16).

Using the Bray-Curtis distance metric, there was a significant
difference across studies in the bacterial community of fecal samples
between carcinoma and controls but not between adenoma and controls
{[}Table S1 \& S2{]}. Many studies with unmatched tissue samples had a
significant difference for both adenoma and carcinoma versus controls
{[}Table S3 \& S4{]} while studies with matched tissue samples had no
differences {[}Table S3 \& S4{]}.

\textbf{\emph{Genera Previously Associated with Carcinoma Predominately
Increases Relative Risk for Carcinoma but not Adenoma:}} The majority of
CRC associated genera for both feces and tissue had a significantly
increased relative risk for carcinoma but not for adenoma {[}Figure
3{]}. In fecal samples the relative risk due to CRC associated genera
was greater than either evenness or Shannon diversity {[}Figure 2 \&
3{]}. Additionally, the relative risk of carcinoma continuously
increased as individuals tested positive for more CRC associated genera
{[}Figure 3B \& 3D{]}. The relative risk effect size was greater for
stool (RR range = 1.61 - 2.74) then for tissue (RR range = 1.21 - 1.81).
This decrease may be explained by the fact that the tissue analysis
include matched samples.

@@

\textbf{\emph{Using the Whole Community Increases Model AUC over CRC
Associated Genera:}} For both fecal and tissue samples (matched and
unmatched) the AUC decreases when only OTUs from the CRC associated
genera are used {[}Figure 4 \& 5{]}. This decrease is observed in both
adenoma and carcinoma groups {[}Figure 4 \& 5{]}. The genus models
generally had similar trends as observed for the OTU based models with
the full genera models performing better then the CRC associated genera
models {[}Figure S2-S3{]}. Both genus models perform similarily in their
ability to be able to predict lesion (adenoma or carcinoma) with
carcinoma having a higher AUC then adenoma {[}Figure S4-S5{]}. Matched
tissue samples for those with carcinoma had an AUC that was more similar
to the adenoma models {[}Figure S4A, S5B, \& S6{]} then carcinoma models
{[}Figure S4B \& S5A{]}.

\textbf{\emph{Majority of Studies are Underpowered for Detecting Small
Effect Size Differences:}} When assessing the power of each study at
different effect sizes the majority of studies for both adenoma and
carcinoma have an 80\% power to detect a 30\% difference {[}Figure 6A \&
B{]}. No single study that was analyzed had the standard 80\% power to
detect an effect size difference that was eqaul to or below 10\%
{[}Figure 6A \& B{]}. In order to achieve adequate power for small
effect sizes it would be necessary to recruit over 1000 individuals for
each arm of the study {[}Figure 6C{]}.
